A Monte Carlo Synthetic-Acceleration Method for Solving the Thermal Radiation Diffusion Equation

### 2012-08-10 doc/jcomp2012/jcomp2012.tex
jcomp updates

    We present a synthetic-acceleration based Monte Carlo method for
    solving the 1T thermal radiation diffusion equation in three
    dimensions. The algorithm performance is compared against
    traditional solution schemes using a Marshak benchmark problem and
    a more sophisticated multiple material problem. Our results show
    that not only can our Monte Carlo method be an effective
    solver for sparse matrix systems, but also that it is competitive
    with deterministic methods including preconditioned Conjugate
    Gradient while producing numerically identical results. We also
    discuss various aspects of preconditioning the method and its
    general applicability to broader classes of problems.

@@

In previous work~\cite{evans_2003}, we developed a residual Monte
Carlo method that was an application of Halton's Sequential Monte
Carlo method~\cite{halton_1962,halton_1994} to the one-dimensional
equilibrium (1T) thermal radiation diffusion equation.  At that time,
Halton's method was not widely known in computational physics.  While
extending our method to three dimensions, we have discovered that the Sequential
Monte Carlo method is actually a variant of iterative refinement cast
as a residual method.  Using Monte Carlo to accelerate a fixed-point
(Richardson) iteration, our new method surpasses our previous 1D
method and allows for extension to more general solution techniques.

The purpose of this study is to demonstrate an efficient Monte Carlo
solution method for 3D, time-dependent, discrete systems. In this
work, we extend our previous work in the following manner:

\begin{itemize}
\item We develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems.
\item We apply this method to the 1T thermal radiation diffusion
  equation in 3D.
\end{itemize}

@@

In order to address the convergence issues plaguing Monte Carlo
solvers, Halton \cite{halton_1962,halton_1994} proposed a Sequential
Monte Carlo method.  This algorithm demonstrated dramatically improved
convergence over regular Monte Carlo.  Nonetheless, this method has
not gained widespread use in computational physics.  In the next
sections, we will briefly describe the Monte Carlo methods that can be
used to solve linear systems and expand them to present our new
synthetic acceleration method.

%%---------------------------------------------------------------------------%%
\subsection{Monte Carlo Matrix Solution Methods}
\label{sec:background}
To establish the mathematical framework for Monte Carlo linear
solvers, we consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\ &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
When the spectral radius of the iteration matrix ($\vH$) is less than
1, we can expand $\vA^{-1}$ using the \latin{Neumann Series},
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} = \sum_{k=0}^{\infty} \vH^k\:.
\end{equation} 
Thus, when $\rho(\vH) < 1$ we can recast the solution vector $\vx$ as
a series,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\ &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb
    + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
With this knowledge in hand, we can write an iterative method that
solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration),
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge for all $\vb \in
\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the
Monte Carlo methods that are described in \S\S~\ref{sec:direct-method}
through \ref{sec:iter-refin-monte} rely on estimating the terms in
Eq.~(\ref{eq:neumann_series}) through random walks.

@@

Similar to the direct method, the adjoint method random walk process
requires a terminating condition.  In all of the work that follows we
utilize a relative weight cutoff.  The weight cutoff is defined as a
fraction of the starting weight such that
\begin{equation}
  W_f = W_c\hat{b}_{i_0}\:,
  \label{eq:weight_cutoff}
\end{equation}
where $W_f$ is the terminating weight of the random walk and $W_c$ is
the input relative weight cutoff. The random walk terminates on the
$m^\text{th}$ step if $W_m < W_f$.

@@

The following scheme will then converge in one iteration:
\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\ \vA \delta \vx^{l+1} =
    \vr^{l+1}\;,
    \label{eq:synthetic-residual-solve}\\ 
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
To accelerate the fixed-point method and avoid inverting $\vA$
directly in Eq.~(\ref{eq:synthetic-residual-solve}), we can instead
use the above ideas to create a new iterative scheme by using the
Monte Carlo method to estimate the solution error. With this in hand,
the Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method can
be defined as follows:
\begin{subequations}
  \label{eq:MCSA-iteration}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\  \vr^{l+1/2} = \vb -
    \vA\vx^{l+1/2}\:,
    \label{eq:MCSA-residual}\\     
     \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\ 
     \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
\end{subequations}
The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that
the Monte Carlo solution only approximately inverts this operator.
Thus, we have defined a scheme in which the initial estimate of $\vx$
in each iteration is updated using a single fixed-point
iteration\footnote{Although, a subspace method could be used instead of
  a stationary method here.}.  The residual is calculated and is used
as a source to estimate the error, $\delta\vx^{l+1/2}$, by solving
Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
method. The error is then used to calculate an updated iterate of the
solution vector, $\vx^{l+1}$.  The entire sequence is iterated to
convergence based on the following stopping
criterion~\cite{kelley_1995},
\begin{equation}
  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

The MCSA method defined in Eq.(~\ref{eq:MCSA-residual_solve}) uses the
adjoint method to estimate the error instead of the direct method
outlined in \S~\ref{sec:direct-method}. To demonstrate the
effectiveness of the adjoint method over the direct method within the
context of MCSA, we choose the two-dimensional time-dependent Poisson
equation as a simple model problem:

@@

@@

\begin{itemize}
\item It reduces the spectral radius such that
  $\rho(\ve{M}^{-1}\ve{D}) < \rho(\ve{D})$ and
  $\rho(\ve{M}^{-1}\ve{D}) < 1$.  Thus, convergence is guaranteed and
  fewer iterations are required resulting in accelerated convergence.
\item The diagonal elements of the preconditioned iteration matrix are
  zero, $\diag(\vI - \ve{M}^{-1}\ve{D}) = 0$.  Thus, there will be no
  computations wasted in transitions $i\rightarrow i$ during the
  random walk process.
\end{itemize}

@@

Here we derive the equations required to solve
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) using the MCSA
method.  As discussed in \S~\ref{sec:iter-refin-monte}, the MCSA
method requires two solution s.pdf:

@@

However, a more natural interpretation for Monte Carlo transport
practitioners is to use the machinery described in
\S~\ref{sec:monte-carlo-matrix} to define Probability Distribution
Functions (PDF) that describe the transport of particles between
cells.  We then do a Monte Carlo transport calculation using
Eq.~(\ref{eq:adjoint-weight}) to calculate the weight change at each
collision. Equation~(\ref{eq:adjoint-probability}) gives the
probability of scattering to an adjacent cell. We tally the
contribution of $\phi$ in each cell using
Eq.~(\ref{eq:adjoint-tally}).  The transport is terminated using a
relative weight cutoff that is defined in
Eq.~(\ref{eq:weight_cutoff}).  We now examine each of these s.pdf in
more detail.

>>>>

    We present a novel synthetic-acceleration based Monte Carlo method
    for solving the equilibrium thermal radiation diffusion equation
    in three dimensions. The algorithm performance is compared against
    traditional solution techniques using a Marshak benchmark problem
    and a more complex multiple material problem. Our results show
    that not only can our Monte Carlo method be an effective solver
    for sparse matrix systems, but also that it performs competitively
    with deterministic methods including preconditioned Conjugate
    Gradient while producing numerically identical results. We also
    discuss various aspects of preconditioning the method and its
    general applicability to broader classes of problems.

@@

In previous work~\cite{evans_2003}, we developed a residual Monte
Carlo method that was an application of Halton's Sequential Monte
Carlo method~\cite{halton_1962,halton_1994} to the one-dimensional
equilibrium (1T) thermal radiation diffusion equation.  At that time,
Halton's method was not widely known in computational physics.  While
extending our method to three dimensions, we have discovered that the
Sequential Monte Carlo method is actually a variant of iterative
refinement cast as a residual method.  Using Monte Carlo to accelerate
a fixed-point (Richardson) iteration, our new method surpasses our
previous one dimensional method and allows for extension to more
general solution techniques.

The purpose of this study is to demonstrate an efficient Monte Carlo
solution method for three dimensional (3D), time-dependent, discrete
systems. In this work, we extend our previous work in the following
manner:

\begin{itemize}
\item We develop a rigorous synthetic-acceleration Monte Carlo method
  for solving sparse matrix systems.
\item We apply this method to the 1T thermal radiation diffusion
  equation in three dimensions.
\item We compare numerical and performance results of the Monte Carlo
  method against traditional deterministic solution methods.
\end{itemize}

@@

In order to address the convergence issues plaguing Monte Carlo
solvers, Halton \cite{halton_1962,halton_1994} proposed a Sequential
Monte Carlo method.  This algorithm demonstrated dramatically improved
convergence over regular Monte Carlo.  Nonetheless, Halton's method
has not gained widespread use in computational physics.  In the next
sections, we will briefly describe the Monte Carlo methods that can be
used to solve linear systems and expand them to present our new
synthetic acceleration method.

%%---------------------------------------------------------------------------%%
\subsection{Monte Carlo Matrix Solution Methods}
\label{sec:background}
To establish the mathematical framework for Monte Carlo linear
solvers, we consider the following matrix equation:
\begin{equation}
  \vA\vx = \vb\:,
  \label{eq:Ax=b}
\end{equation}
which can be written,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vA)\vx + \vb\\ &= \vH\vx + \vb\:.
  \end{split}
  \label{eq:point_iteration}
\end{equation}
When the spectral radius ($\rho$) of the iteration matrix ($\vH$) is
less than 1, we can expand $\vA^{-1}$ using the \latin{Neumann
  Series},
\begin{equation}
  \vA^{-1} = (\vI - \vH)^{-1} = \sum_{k=0}^{\infty} \vH^k\:.
\end{equation} 
Thus, when $\rho(\vH) < 1$ we can recast the solution vector $\vx$ as
a series,
\begin{equation}
  \begin{split}
    \vx &= (\vI - \vH)^{-1}\vb\\ &= \vb + \vH\vb + \vH^2\vb + \vH^3\vb
    + \ldots\:.
    \label{eq:neumann_series}
  \end{split}
\end{equation}
With this knowledge in hand, we can write an iterative method that
solves Eq.~(\ref{eq:Ax=b}) (Richardson's Iteration),
\begin{equation}
  \vx^{k+1} = \vH\vx^k + \vb\:.
  \label{eq:richardson}
\end{equation}
Equation~(\ref{eq:richardson}) will converge for all $\vb \in
\mathcal{R}^{N}$ when $\rho(\vH) < 1$~\cite{kelley_1995}.  All of the
Monte Carlo methods that are described in \S\S~\ref{sec:direct-method}
through \ref{sec:iter-refin-monte} rely on estimating the terms in
Eq.~(\ref{eq:neumann_series}) through random walks.

@@

Similar to the direct method, the adjoint method random walk process
requires a terminating condition.  In all of the work that follows we
utilize a relative weight cutoff.  The relative weight cutoff is
defined as a fraction of the starting weight such that
\begin{equation}
  W_f = W_c\hat{b}_{i_0}\:,
  \label{eq:weight_cutoff}
\end{equation}
where $W_f$ is the terminating weight of the random walk and $W_c$ is
the input relative weight cutoff. The random walk terminates on the
$m^\text{th}$ step if $W_m < W_f$.

@@

The following scheme will then converge in one iteration:
\begin{subequations}
  \begin{gather}
    \vx^{l+1} = (\vI - \vA)\vx^l + \vb\:,\\ \vA \delta \vx^{l+1} =
    \vr^{l+1}\;,
    \label{eq:synthetic-residual-solve}\\ 
    \vx=\vx^{l+1}+\delta\vx^{l+1}\:.
  \end{gather}
\end{subequations}
To accelerate the fixed-point method and avoid inverting $\vA$
directly in Eq.~(\ref{eq:synthetic-residual-solve}), we can instead
use the above ideas to create a new iterative scheme by using the
Monte Carlo method to estimate the solution error. With this in hand,
the Fixed-Point Monte Carlo Synthetic-Acceleration (MCSA) method can
be defined as follows:
\begin{subequations}
  \label{eq:MCSA-iteration}
  \begin{gather}
    \vx^{l+1/2} = (\vI - \vA)\vx^l + \vb\:,\\  \vr^{l+1/2} = \vb -
    \vA\vx^{l+1/2}\:,
    \label{eq:MCSA-residual}\\     
     \hat{\vA}\delta\vx^{l+1/2} = \vr^{l+1/2}\:,
    \label{eq:MCSA-residual_solve}\\ 
     \vx^{l+1}=\vx^{l+1/2}+\delta\vx^{l+1/2}\:.
  \end{gather}
\end{subequations}
The hat on $\vA$ in Eq.~(\ref{eq:MCSA-residual_solve}) indicates that
the Monte Carlo solution only approximately inverts this operator.
Thus, we have defined a scheme in which the initial estimate of $\vx$
in each iteration is updated using a single fixed-point
iteration\footnote{Although, a subspace method could be used instead
  of a stationary method here.}.  The residual is calculated and used
as a source to estimate the error, $\delta\vx^{l+1/2}$, by solving
Eq.~(\ref{eq:MCSA-residual_solve}) via the Monte Carlo adjoint
method. The error is then used to calculate an updated iterate of the
solution vector, $\vx^{l+1}$.  The entire sequence is iterated to
convergence based on the following stopping
criterion~\cite{kelley_1995},
\begin{equation}
  \|\vr\|_{\infty} < \epsilon\cdot\|\vb\|_\infty\:.
  \label{eq:stopping-criteria}
\end{equation}

@@

The MCSA method defined in Eq.~(\ref{eq:MCSA-iteration}) uses the
adjoint method to estimate the error in
Eq.~(\ref{eq:MCSA-residual_solve}) instead of the direct method
outlined in \S~\ref{sec:direct-method}. To demonstrate the
effectiveness of the adjoint method over the direct method within the
context of MCSA, we choose the two-dimensional time-dependent Poisson
equation as a simple model problem:

@@

Using the mathematical preliminaries developed in the previous
section, we can now generate a discrete form of the radiation
diffusion equation. We will show that this discrete form is amenable
to interpretation using Monte Carlo concepts and can be written in a
way that a random walk sequence can be generated.

@@

\begin{itemize}
\item It reduces the spectral radius such that
  $\rho(\ve{M}^{-1}\ve{D}) < \rho(\ve{D})$ and
  $\rho(\ve{M}^{-1}\ve{D}) < 1$.  Thus, convergence is guaranteed and
  fewer iterations are required resulting in accelerated convergence.
\item The diagonal elements of the preconditioned iteration matrix are
  zero, $\diag(\vI - \ve{M}^{-1}\ve{D}) = 0$.  Thus, there will be no
  computations wasted for in-state transitions $i\rightarrow i$ during
  the random walk process.
\end{itemize}

@@

Here we derive the equations required to solve
Eq.~(\ref{eq:preconditioned-discrete-diffusion}) using the MCSA
method.  As discussed in \S~\ref{sec:iter-refin-monte}, the MCSA
method requires two solutions:

@@

However, a more natural interpretation for Monte Carlo transport
practitioners is to use the machinery described in
\S~\ref{sec:monte-carlo-matrix} to define Probability Distribution
Functions (PDF) that describe the transport of particles between
cells.  We then do a Monte Carlo transport calculation using
Eq.~(\ref{eq:adjoint-weight}) to calculate the weight change at each
collision. Equation~(\ref{eq:adjoint-probability}) gives the
probability of scattering to an adjacent cell. We tally the
contribution of $\phi$ in each cell using
Eq.~(\ref{eq:adjoint-tally}).  The transport is terminated using a
relative weight cutoff that is defined in
Eq.~(\ref{eq:weight_cutoff}).  We now examine each of these in
more detail.
